# Title

### STAT 420 Final Data Analysis Project - Team KWS

### 

# Introduction

Life expectancy varies greatly across the world. Understanding the factors that influence life expectancy can provide insights for policymakers, healthcare providers, and individuals. This project aims to delve into the determinants of life expectancy.

The dataset used is from the World Health Organization (WHO) and United Nations, which provides a wealth of data related to life expectancy across 193 countries. The dataset, publicly available on Kaggle under the title "Life Expectancy (WHO)", comprises 22 columns and 2,938 observations. Each observation represents the state of a given country in a particular year, with 20 predictor variables providing comprehensive information about various factors potentially impacting life expectancy.

The predictor variables span across a broad spectrum, including immunization-related factors, mortality factors, economic indicators, and social factors. Among these, we believe certain variables like the status of the country (developed vs developing), infant mortality rate, adult mortality rate, health expenditure, GDP, and average years of schooling may have significant importance in predicting life expectancy.

Our objective in this project is to create a model that can effectively predict life expectancy based on these predictors. We're particularly interested in this dataset due to our curiosity about the factors influencing life expectancy and our desire to uncover the differences between countries that lead to a wide range of life expectancies. By building this model, we aim to provide insights that could potentially guide health policy and contribute to improvements in global health outcomes.

# Methods

### Loading Data

```{r}
#read data
library(readr)
life_expectancy_data_full = read_csv("Life Expectancy Data.csv")
```

We will remove rows with empty values to make analysis easier.

```{r}
#remove rows with empty vals
life_expectancy_data = na.omit(life_expectancy_data_full)
View(life_expectancy_data)
```


For our analysis, we have chosen to convert the predictors Status and Country into factor variables, so that we can treat them as categories.

```{r}
#convert Status (levels: Developing, Developed) to factor variable
life_expectancy_data$Status = as.factor(life_expectancy_data$Status)

#convert Country to factor variable
life_expectancy_data$Country = as.factor(life_expectancy_data$Country)

```

### Exploratory Analysis

The pairs plot of all of the variables was very large, so we will display correlations between the numeric variables in the dataset with a Correlogram instead.

```{r}
#pairs(life_expectancy_data)

#Pairs plot is very large, so we create a Correlogram to view correlations 


#find pairwise correlations between all numeric variables in dataset
numeric_cols =  unlist(lapply(life_expectancy_data, is.numeric))
life_expectancy_data_numeric = life_expectancy_data[ , numeric_cols]
View(life_expectancy_data_numeric)

correlations = cor(life_expectancy_data_numeric)

#plot correlogram (positive correlations are blue, negative correlations are red)
library(corrplot)

corrplot(correlations)
```

We notice that there are a few variables with very high correlations, so we will attempt to address this issue in our modeling.

### Initial Additive Model

First, we create a main effects model using all of the available predictors in the dataset to predict life expectancy.

```{r}
#create initial additive model with all predictors

additive_model = lm(life_expectancy_data$`Life expectancy` ~ ., data = life_expectancy_data)
#summary(additive_model)
```

### Analyze Model

```{r}

#Calculate LOOCV RMSE
calc_loocv_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}

# Function to calculate model metrics

calc_model_metrics = function(model) {
  # Calculate adjusted R squared
  model_adjr2 = summary(model)$adj.r.squared
  
  # Calculate RMSE
  model_rmse = sqrt(mean(resid(model) ^ 2))
  
  # Calculate LOOCV RMSE
  model_loocv_rmse = sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
  
  # Calculate AIC
  model_AIC = extractAIC(model)[2] # get only the AIC value, not the degrees of freedom
  
  # Return a list containing the calculated metrics
  list(Adj_R_Squared = model_adjr2, RMSE = model_rmse, LOOCV_RMSE = model_loocv_rmse, AIC = model_AIC)
}

additive_model_metrics = as.data.frame(calc_model_metrics(additive_model))

library(knitr)
knitr::kable(additive_model_metrics, caption = "Additive Model Summary", digits = 4)
```

### Check model assumptions

```{r}
#------- Check model assumptions -------------

#Fitted vs Residuals Plot

plot(fitted(additive_model), resid(additive_model), col = "grey", pch = 20,
     xlab = "Fitted", ylab = "Residuals", main = "Data from Model 1")
abline(h = 0, col = "darkorange", lwd = 2)

#Q-Q Plot

qqnorm(resid(additive_model), main = "Normal Q-Q Plot, fit_1", col = "darkgrey")
qqline(resid(additive_model), col = "dodgerblue", lwd = 2)

#confirm issues via more formal tests

library(lmtest)
bptest(additive_model)
#result shows data violates constant variance assumption!

shapiro.test(resid(additive_model))
#result shows data violates normality assumption!
```

```{r}
#---------- Next step -------------------------------

#We will attempt to fix these issues by removing highly correlated predictors
#under 5 death & infant deaths : cor = 0.997
#thinness 1-19 years & thinness 5-9 years : cor = 0.928
#We'll remove one from each of the above pairs of predictors and reassess model

#Removing infant deaths and thinness 1-19 years
corr_removed_data_1 = life_expectancy_data[ , !(names(life_expectancy_data) %in% c('infant deaths', 'thinness 1-19 years'))]
corr_removed_model_1 = lm(`Life expectancy` ~ ., data = corr_removed_data_1)
corr_removed_metrics_1 = calc_model_metrics(corr_removed_model_1)


#Removing infant deaths and thinness 5-9 years
corr_removed_data_2 = life_expectancy_data[ , !(names(life_expectancy_data) %in% c('infant deaths', 'thinness 5-9 years'))]
corr_removed_model_2 = lm(`Life expectancy` ~ ., data = corr_removed_data_2)
corr_removed_metrics_2 = calc_model_metrics(corr_removed_model_2)


#Removing under 5 deaths and thinness 1-19 years
corr_removed_data_3 = life_expectancy_data[ , !(names(life_expectancy_data) %in% c('under 5 death', 'thinness 1-19 years'))]
corr_removed_model_3 = lm(`Life expectancy` ~ ., data = corr_removed_data_3)
corr_removed_metrics_3 = calc_model_metrics(corr_removed_model_3)

#Removing under 5 deaths and thinness 5-9 years
corr_removed_data_4 = life_expectancy_data[ , !(names(life_expectancy_data) %in% c('under 5 death', 'thinness 5-9 years'))]
corr_removed_model_4 = lm(`Life expectancy` ~ ., data = corr_removed_data_4)
corr_removed_metrics_4 = calc_model_metrics(corr_removed_model_4)

# Create a data frame summarizing the model metrics
model_summary = data.frame(
  Variables_Removed = c("infant deaths + thinness 1-19", "infant deaths + thinness 5-9", "under 5 deaths + thinness 1-19", "under 5 deaths + thinness 5-9 years"),
  Adjusted_R_Squared = c(corr_removed_metrics_1$Adj_R_Squared, corr_removed_metrics_2$Adj_R_Squared, corr_removed_metrics_3$Adj_R_Squared, corr_removed_metrics_4$Adj_R_Squared),
  RMSE = c(corr_removed_metrics_1$RMSE, corr_removed_metrics_2$RMSE, corr_removed_metrics_3$RMSE, corr_removed_metrics_4$RMSE),
  LOOCV_RMSE = c(corr_removed_metrics_1$LOOCV_RMSE, corr_removed_metrics_2$LOOCV_RMSE, corr_removed_metrics_3$LOOCV_RMSE, corr_removed_metrics_4$LOOCV_RMSE),
  AIC = c(corr_removed_metrics_1$AIC, corr_removed_metrics_2$AIC, corr_removed_metrics_3$AIC, corr_removed_metrics_4$AIC)
)

library(knitr)
knitr::kable(model_summary, caption = "Model Summary", digits = 4)
# removing under 5 deaths + thinness 1-19 is the best option



#---------- Next step -------------------------------
#We will attempt to further improve the model by log transforming certain predictor variables (Population & GDP??)

log_transform_1 = lm(`Life expectancy` ~ . + log(Population) + log(GDP), data = life_expectancy_data)
log_transformed_metrics_1 = calc_model_metrics(log_transform_1)

log_transform_2 = lm(`Life expectancy` ~ . + log(GDP), data = life_expectancy_data)
log_transformed_metrics_2 = calc_model_metrics(log_transform_2)

log_transform_3 = lm(`Life expectancy` ~ . + log(Population), data = life_expectancy_data)
log_transformed_metrics_3 = calc_model_metrics(log_transform_3)


log_model_summary = data.frame(
  Variables_Log_Transformed = c("Population + GDP", "GDP", "Population"),
  Adjusted_R_Squared = c(log_transformed_metrics_1$Adj_R_Squared, log_transformed_metrics_2$Adj_R_Squared, log_transformed_metrics_3$Adj_R_Squared),
  RMSE = c(log_transformed_metrics_1$RMSE, log_transformed_metrics_2$RMSE, log_transformed_metrics_3$RMSE),
  LOOCV_RMSE = c(log_transformed_metrics_1$LOOCV_RMSE, log_transformed_metrics_2$LOOCV_RMSE, log_transformed_metrics_3$LOOCV_RMSE),
  AIC = c(log_transformed_metrics_1$AIC, log_transformed_metrics_2$AIC, log_transformed_metrics_3$AIC)
)

knitr::kable(log_model_summary, caption = "Log Transformed Model Summary", digits = 4)

#log(GDP) is the best result, although it is only a very minimal improvement



#---------- Next step -------------------------------
#We will attempt to further improve the model by log transforming certain predictor variables (Population & GDP??)



#---------- Next step --------------------------------

#We will use AIC and BIC backwards search to attempt to further improve the model
```

# Results

# Discussions

# Appendix

# Saitejas Mopuri, Kimberly Martin, Woo Jeon
